### 2024-09-25 paper/paper.qmd
Initial code for paper code

---
title: "Annual Analysis of Delays in TTC Streetcars for the year 2023"
author: 
  - Aakash Vaithyanathan
thanks: "Code and data are available at: https://github.com/aakash2002/study_of_ttc_streetcar_delays"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

data <- read_csv(here::here("data/analysis_data/cleaned_ttc_streetcar_delay_2023.csv"))
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false
#| label: fig-delay-by-line
#| fig-cap: Average annual delay in minutes for each streetcar line

# Grouping and summarizing the data
grp_data <- data |> 
  group_by(line) |>
  summarise(average_delay = mean(min_delay), .groups = 'drop')

# Create the bar plot
ggplot(grp_data, aes(x = factor(line), y = average_delay)) +
  geom_bar(stat = "identity", width = 0.7, fill = 'lightblue') +
  geom_text(aes(label = round(average_delay, 2)), 
            vjust = -0.5,  # Position the text above the bar
            size = 2) +  # Adjust the size of the text
  ggtitle("Bar plot of average delay in minutes by line") +
  xlab("Streetcar Line") +
  ylab("Average Delay in Minutes") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        title = element_text(hjust = 0.5))

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-delay-for-day-shift
#| fig-cap: Average delay in minutes day shift
#| echo: false
#| warning: false
#| message: false

# Filter the data for times between 06:00 and 18:00 i.e day shift
avg_delay_day_shift <- data |> 
  group_by(hour) |>
  summarise(average_delay = mean(min_delay), .groups = 'drop') |>
  filter(hour > "06" & hour < "19")

dayshift <- ggplot(avg_delay_day_shift, aes(x = hour, y = average_delay)) +
    geom_line(color = "red", size = 1) +
    geom_point(color = "blue", size = 2) +
    ggtitle("Average delay in minutes for morning shift") +
    xlab("Hour of Day") +
    ylab("Average Delay (Minutes)") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
    )

# Filter the data for times between 19:00 and 00:00 i.e night shift
avg_delay_night_shift <- data |> 
  group_by(hour) |>
  summarise(average_delay = mean(min_delay), .groups = 'drop') |>
  filter(hour >= "19" | hour <= "01")

nightshift <- ggplot(avg_delay_night_shift, aes(x = hour, y = average_delay)) +
    geom_line(color = "blue", size = 1) +
    geom_point(color = "red", size = 2) +
    ggtitle("Average delay in minutes for night shift") +
    xlab("Hour of Day") +
    ylab("Average Delay (Minutes)") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
    )

dayshift
nightshift
```

Talk way more about it.

```{r}
#| label: fig-freq-delay-incidents
#| fig-cap: Frequency plot for cause of streetcar delay by incidents
#| echo: false
#| warning: false
#| message: false

freq_incidence <- data |> 
  group_by(incident) |> 
  summarise(frequency = n(), .groups = 'drop')

ggplot(freq_incidence, aes(x = incident, y = frequency)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  geom_text(aes(label = frequency), 
            vjust = -0.5,  # Position the text above the bar
            size = 2) +  # Adjust the size of the text
  ggtitle("Frequency of each delay incident in the year 2023") +
  xlab("Delay Incident Type") +
  ylab("Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center title
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
  )
```

```{r}
#| label: fig-min-delay-by-season
#| fig-cap: Average delay in minutes by seasons in a yea
#| echo: false
#| warning: false
#| message: false

delay_by_season <- data |>
  group_by(season) |>
  summarise(average_delay = mean(min_delay), .groups = 'drop')

ggplot(delay_by_season, aes(x = season, y = average_delay)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  ggtitle("Average delay in minutes by seasons in a year") +
  xlab("Type of Season") +
  ylab("Average delay in minutes") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center title
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
  )
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

\newpage

# References

### 2024-09-25 scripts/00-simulate_data.R

@@ -19,6 +19,8 @@ end_date <- as.Date("2023-12-31")
 
 # Set the number of random dates you want to generate
 number_of_dates <- 100
+from_stations <- c("Finch", "Sheppard Yongue", "Lawrence Station", "Bloor Station")
+to_stations <- c("St George Station", "Union Station", "Queen Station", "Dundas West", "Museum")
 
 data <-
   tibble(
@@ -35,13 +37,13 @@ data <-
       c("Traffic", "Emergency", "Weather", "Route Change"), 
       size = number_of_dates, replace = TRUE
       ),
-    from_station = sample(
-      c("Finch", "Sheppard Yongue", "Lawrence Station", "Bloor Station"), 
-      size = number_of_dates, replace = TRUE
+    from_station = sample(from_stations, 
+                          size = number_of_dates, 
+                          replace = TRUE
     ),
-    to_station = sample(
-      c("St George Station", "Union Station", "Queen Station", "Dundas West", "Museum"), 
-      size = number_of_dates, replace = TRUE
+    to_station = sample(to_stations, 
+                        size = number_of_dates, 
+                        replace = TRUE
     ),
   )
 

### 2024-09-25 scripts/02-data_cleaning.R

@@ -20,11 +20,23 @@ cleaned_data <-
   separate(col = date,
            into = c("year", "month"),
            sep = "-") |>
-  tidyr::drop_na() |>
+  separate(col = time,
+           into = c("hour", "minute"),
+           sep = ":") |>
+  mutate(season = case_when(
+    month %in% c("03", "04", "05") ~ "Spring",  # March, April, May
+    month %in% c("06", "07", "08") ~ "Summer",  # June, July, August
+    month %in% c("09", "10", "11") ~ "Fall",     # September, October, November
+    month %in% c("12", "01", "02") ~ "Winter"    # December, January, February
+  )) |>
+  drop_na() |>  # Ensure drop_na() is applied to the cleaned_data
   filter(min_delay != 0) |>
   filter(min_gap != 0) |>
-  mutate(time = substr(time, 1, 5)) |>
-  select(year, month, day, location, incident, min_delay, line)
+  select(year, month, hour, minute, incident, min_delay, line, season)
+
+# Since the delay time for some dates are over many hours, we will restrict
+# ourselves to at most 2 hour delays in our study
+cleaned_data <- cleaned_data |> filter(min_delay < 121)
 
 # Check if any column contains NA data or not
 any(is.na(cleaned_data))

### 2024-09-25 paper/paper.qmd
Added introduction and details about graphs

@@
 title: "Annual Analysis of Delays in TTC Streetcars for the year 2023"
+subtitle: "Is there a way to make commuter life easier?"
 author: 
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Toronto Transit Commission (TTC) is the biggest and most used transit authority in Toronto.
+It was established on September 1, 1921 and since then supports various means of ground transportation like streetcar, shuttle bus and the subway system.
+Despite being the most used transit system, it does come with several challenges one of the most notable one being the delays experienced between the various stops.
+One such transportation system we focus on in this paper is streetcars.
 
-The remainder of this paper is structured as follows. @sec-data....
+These delays make it especially difficult for individuals that commute on a daily basis.
+Furthermore, one can image the inconveniences caused due to extended delays by streetcars during weather conditions like the winter months.
+In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced.
+We make use of the dataset made available by @opendatatoronto for the year 2023.
 
-# Data {#sec-data}
+The remainder of this paper is structured as follows.
+We first give a brief overview of the dataset under @dataset-desc.
+Then, we present some graphs on the results extracted from our collected data found under @results.
+Lastly, we conclude with a discussion of our results and any limitations in our approach found under @discussion.
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+# Dataset Description {#dataset-desc}
 
+bla bla bla
+
+# Results {#results}
+## Average Delay By Streetcar Line
 ```{r}
@@
 
-Talk more about it.
+@fig-delay-by-line highlights the annual average delay in minutes for every streetcar line for the year 2023.
+The 300-series lines also called the Blue Network Lines (cite) are the nighttime streetcar lines that start at 1:30 AM.
+These buses average the most delays reaching at most 30 minutes for line 312.
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+The 500-series lines are the daytime lines that run from 8 AM. These buses average between 6 to 23.6 minutes with the longest delay experienced by line 507.
 
+## Average Delay By Working Shift
+### Delays During Dayshift
 ```{r}
-#| label: fig-delay-for-day-shift
-#| fig-cap: Average delay in minutes day shift
+#| label: fig-delay-dayshift
+#| fig-cap: Average delay in minutes for dayshift
 #| echo: false
@@
 
-# Filter the data for times between 06:00 and 18:00 i.e day shift
+# Filter the data for times between 08:00 and 18:00 i.e day shift
 avg_delay_day_shift <- data |> 
@@
   summarise(average_delay = mean(min_delay), .groups = 'drop') |>
-  filter(hour > "06" & hour < "19")
+  filter(hour >= "08" & hour <= "18")
 
-dayshift <- ggplot(avg_delay_day_shift, aes(x = hour, y = average_delay)) +
+ggplot(avg_delay_day_shift, aes(x = hour, y = average_delay)) +
   geom_line(color = "red", size = 1) +
   geom_point(color = "blue", size = 2) +
+  geom_text(aes(label = round(average_delay, 2)), 
+            vjust = -1, 
+            size = 2) + 
   ggtitle("Average delay in minutes for morning shift") +
@@
   )
+```
+
+@fig-delay-dayshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the day.
+We observe that there is high variability in the delays throughout the day with the minimum delay being about 13.78 minutes and the maximum being roughly 20 minutes.
+
+### Delays During Nightshift
+```{r}
+#| label: fig-delay-nightshift
+#| fig-cap: Average delay in minutes for nightshift
+#| echo: false
+#| warning: false
+#| message: false
 
@@
 
-nightshift <- ggplot(avg_delay_night_shift, aes(x = hour, y = average_delay)) +
+# Specify the desired order for the hours (military time format)
+hour_order <- c("17", "18", "19", "20", "21", "22", "23", "00", "01")
+
+# Convert the 'hour' column to a factor with the specified order
+avg_delay_night_shift$hour <- factor(avg_delay_night_shift$hour, levels = hour_order)
+
+# Plot the graph with the reordered x-axis
+ggplot(avg_delay_night_shift, aes(x = hour, y = average_delay)) +
   geom_line(color = "blue", size = 1) +
   geom_point(color = "red", size = 2) +
+  geom_text(aes(label = round(average_delay, 2)), 
+            vjust = -1, 
+            size = 2) + 
   ggtitle("Average delay in minutes for night shift") +
@@
   )
-
-dayshift
-nightshift
 ```
 
-Talk way more about it.
+@fig-delay-nightshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the night We observe that there is low variability in the delays throughout the night with the minimum delay being about 16.62 minutes and the maximum roughly 20 minutes.
 
+## Frequency of delays by incidents
 ```{r}
@@
 
-# Discussion
+# Discussion {#discussion}
 

### 2024-09-25 paper/paper.qmd
Added discussion for some of the reported result graphs

@@
             size = 2) +  # Adjust the size of the text
-  ggtitle("Bar plot of average delay in minutes by line") +
+  ggtitle("Average delay in minutes by streetcar line") +
   xlab("Streetcar Line") +
@@
 
-## Average Delay By Working Shift
-### Delays During Dayshift
+## Average Delays During Dayshift
 ```{r}
@@
 
-# Filter the data for times between 08:00 and 18:00 i.e day shift
+# Filter the data for times between 08:00 and 01:00 i.e day shift
 avg_delay_day_shift <- data |> 
@@
   summarise(average_delay = mean(min_delay), .groups = 'drop') |>
-  filter(hour >= "08" & hour <= "18")
+  filter(hour >= "08" | hour <= "01")
 
@@
 ```
-
 @fig-delay-dayshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the day.
@@
 
-### Delays During Nightshift
+## Average Delays During Nightshift
 ```{r}
@@
 
-# Filter the data for times between 19:00 and 00:00 i.e night shift
+# Filter the data for times between 02:00 and 07:00 i.e night shift
 avg_delay_night_shift <- data |> 
   group_by(hour) |>
-  summarise(average_delay = mean(min_delay), .groups = 'drop') |>
-  filter(hour >= "19" | hour <= "01")
-
-# Specify the desired order for the hours (military time format)
-hour_order <- c("17", "18", "19", "20", "21", "22", "23", "00", "01")
-
-# Convert the 'hour' column to a factor with the specified order
-avg_delay_night_shift$hour <- factor(avg_delay_night_shift$hour, levels = hour_order)
+  filter(hour >= "02" & hour <= "07") |>
+  summarise(average_delay = mean(min_delay), .groups = 'drop')
 
@@
 ```
-
-@fig-delay-nightshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the night We observe that there is low variability in the delays throughout the night with the minimum delay being about 16.62 minutes and the maximum roughly 20 minutes.
+@fig-delay-nightshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the night.
+We observe that there is decreasing trend in the delays throughout the night after 3 AM with the minimum delay being about 14 minutes and the maximum roughly 27 minutes.
 
@@
 ```
+@fig-freq-delay-incidents highlights the total annual delays incurred based on incidents reported by TTC for the reason.
+We can see for the year of 2023, operation delays were the most frequently occurring including a total of 2463 instances and rail switches were the least occuring with reported instances being only about 20.
 
+## Delay by Seasons
 ```{r}
-#| label: fig-min-delay-by-season
+#| label: fig-delay-by-season
 #| fig-cap: Average delay in minutes by seasons in a yea
@@
 
+@fig-delay-by-season highlights the annual average delay in minutes based on the seasons.
+We observe that were isn't as much variation between the streetcar delays between the fall, spring, summer and winter months with each averaging around 16 minutes delay.
+
 # Discussion {#discussion}
+The above analysis provide valuable information regarding causes for TTC streetcar delays. Let us take a closer look into each of these.
+## Delays by bus lines
+The results from @fig-delay-by-line shows that blue-network (night shift) buses incur the highest average delay.
+This result makes sense as compared to the day shift buses, there are far fewer night shift buses resulting in fewer drivers for various locations the buses may take.
+Additionally, we also note that the 900-series streetcar line that go to the Airport incur the longest average delay of almost 54 minutes.
 
-## First discussion point {#sec-first-point}
+## Average delay for morning shift
+The results from @fig-delay-dayshift showcases the variability in average delays by the hour during the morning shift working hours observed between 8 AM to 6 PM.
+We observe that the results are in-line with the expected behavior of longer average delays of around 18 minutes during peak hours like 8 AM or 5 PM which is usually when a large working population leave and return from work.
+This can result in higher traffic hours on average causing an expected delay in streetcar arrivals.
+Additionally, we notice that throughout the day there are several points of interest for peak delays as seen at 11 AM or 1 PM incurring an average delay of 19 minutes.
+This observation could indicate the need for travel to various locations like grocery stores, perhaps to a friend's house or classes for university students.
+During these hours, there is heavy traffic and these suggest a potential reason for delays.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
-
-## Second discussion point
+## Average delay for night shift
+The results from @fig-delay-nightshift shows a decreasing trend in the average delays by the hour during the night shift working hours observed between 2 AM to 7 AM.
+These findings are what we might expect as there is fewer traffic delays during the night resulting in a more consistent decreasing delay pattern.
+A possible reason for longer delay duration despite fewer traffic could be the fact there there are far fewer blue-network (night shift) buses that run resulting in fewer drivers that require to cover longer distances.
+The graph shows that as we get later into the night shift (i.e early morning) we notice the delays dropping as low as roughly 14 minutes.
 

### 2024-09-26 paper/paper.qmd
Added discussion, results, limitations, dataset description and references

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
-format: pdf
+abstract: "This paper aims to study the causes of delays in Toronto Transit Commissions (TTC) streetcar service, the variation in delays by the hours in working shifts and whether there are any seasonal relationships to the observed delays for the year 2023.
+By using the data provided by OpenData Toronto, we are able to understand the observed relationships between the several fields of interests like minimum delay, hour, season, streetcar line and incident type.
+Our study finds that there is higher reported delayed times during night shift and higher delays reported during peak hours that fall during the working 9AM - 5PM time period.
+Interestingly we did not observe any variation in delays during the various seasons and understand that to make stronger inferences between the temporal relationship, we need to examine longer historic data."
+format: 
+  pdf: 
+    fig-pos: "H"
+    fig-caption: true
 number-sections: true
 bibliography: references.bib
+link-citations: true
 ---
@@
 library(ggplot2)
+library(gt)
 
 data <- read_csv(here::here("data/analysis_data/cleaned_ttc_streetcar_delay_2023.csv"))
+
+table_styling <- function(gt_tbl) {
+  gt_tbl |> 
+    cols_label(
+      year = "Year",
+      month = "Month",
+      hour = "Hour",
+      incident = "Incident Type",
+      min_delay = "Min Delay (minutes)",
+      line = "Streetcar Line",
+      season = "Season"
+    ) |>
+    tab_options(
+      column_labels.background.color = 'dodgerblue1',
+      table_body.hlines.width = px(0) 
+    )
+  }
 ```
@@
 In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced.
-We make use of the dataset made available by @opendatatoronto for the year 2023.
+We make use R (@citeR) and the dataset made available by OpenData Toronto (@opendatatoronto) for the year 2023.
+The reported graphs in our study is generated using ggplot (@ggplot2) and the tables are generated using the gt (@gt) library.
+Code cleanup is performed using the Janitor (@janitor) package.
 
 The remainder of this paper is structured as follows.
-We first give a brief overview of the dataset under @dataset-desc.
-Then, we present some graphs on the results extracted from our collected data found under @results.
-Lastly, we conclude with a discussion of our results and any limitations in our approach found under @discussion.
+We first give a brief overview of the dataset under [datasetdesc] section.
+Then, we present some graphs on the results extracted from our collected data found under [results] section.
+Lastly, we conclude with a discussion of our results found under [discussion] section, highlight the limitations in our conducted study found under [limitations] section and propose possible future next steps for this project under [nextsteps] section.
 
-# Dataset Description {#dataset-desc}
+# Dataset Description {#datasetdesc}
 
-bla bla bla
+## Overview of dataset
+The following data was made available from OpenData Toronto (@opendatatoronto) for the year 2023.
+The data was extracted from the server and saved locally using in a raw data folder.
+Some of the features part of this dataset are date, time, incident, min delay, bound and location.
+
+## Assumption made in dataset {#assumption}
+We observe that our dataset has several rows where the minimum delay time is over 900 minutes or are equal to 0.
+These values can heavily skew the results that we may observe.
+Therefore, one key assumption we make in our study is that we limit the maximum delay time to upto 2 hours or 120 minutes.
+As a result of this filtering, we only lose about 130 data and still retain about 10723 data entries to use in our study.
+
+## Data preprocessing and cleaning
+The raw data is undergone a series of data cleaning steps to ensure adequate usability for the study.
+We first use the Janitor (@janitor) package to clean the data columns removing any spaces between them and using a consistent naming stype for them.
+Next, we proceed to split the date field into year and month fields and split the time field extracting the hour of day from it.
+Following this, we create a new column to denote the season based on the month of the year from the data.
+This helps during the analysis stage when we need to make inferences about the seasonal variation in average delays in streetcars.
+
+We then proceed to drop the rows with missing values or NA.
+Out of the total 14413 rows of data, 2399 of them contain NA values in total which doesn't heavily impact our data.
+Therefore, no transformations are necessary to populate missing entries in our dataset.
+Lastly, as highlighed in the [assumption] section, we filter our dataset to only include delays of upto 120 minutes.
+
+As a result of the above described data cleaning process, we select our desired features from the dataset. The following table shows the first five rows of our data and the selected features.
+
+```{r}
+#| include: true
+#| warning: false
+#| message: false
+#| echo: false
+
+# Create a table highlighting the first five rows and the selected features in the data
+limited_row_data <- data |> head(5)
+
+table_features <- gt(limited_row_data)
+table_features |> 
+  table_styling() |>
+  tab_header(
+    title = "Selected Features from TTC Streetcar Delay Data"
+    ) |>
+  tab_style(
+    style = cell_text(
+      weight = 'bold',
+      font = google_font('Merriweather')
+    ),
+    locations = cells_title(group = 'title')
+  ) |>
+  tab_style(
+    style = cell_fill(color = 'grey90'),
+    locations = cells_body(rows = seq(1, nrow(limited_row_data), 2))
+  )
+cat("Table 1: First first five rows in the cleaned dataset with selected features")
+
+```
+
+## Measurement of units for selected features
+
+The following table describes the data types for the selected features in our study from the cleaned dataset.
+```{r}
+#| include: true
+#| warning: false
+#| message: false
+#| echo: false
+
+# Get feature names
+features <- colnames(limited_row_data)
+
+# Get data types for each feature
+datatypes <- sapply(limited_row_data, class)
+
+# Combine into a table (data frame)
+table <- data.frame(Feature = features, DataType = datatypes)
+table_feature_datatypes <- gt(table)
+table_feature_datatypes |>
+  tab_header(
+    title = "Selected Features and their Data Types"
+    ) |>
+  tab_style(
+    style = cell_text(
+      weight = 'bold',
+      font = google_font('Merriweather')
+    ),
+    locations = cells_title(group = 'title')
+  ) |>
+  tab_style(
+    style = cell_text(weight = 'bold'),
+    locations = cells_column_labels()
+  ) |>
+  tab_style(
+    style = cell_fill(color = 'grey90'),
+    locations = cells_body(rows = seq(1, nrow(table), 2))
+  )
+  cat("Table 2: Datatypes for the selected features from the cleaned dataset.")
+```
 
@@
 @fig-delay-by-line highlights the annual average delay in minutes for every streetcar line for the year 2023.
-The 300-series lines also called the Blue Network Lines (cite) are the nighttime streetcar lines that start at 1:30 AM.
+The 300-series lines also called the Blue Network Routes (@ttcRoutesSchedule) are the nighttime streetcar lines that start at 1:30 AM.
 These buses average the most delays reaching at most 30 minutes for line 312.
@@
 The above analysis provide valuable information regarding causes for TTC streetcar delays. Let us take a closer look into each of these.
+
 ## Delays by bus lines
-The results from @fig-delay-by-line shows that blue-network (night shift) buses incur the highest average delay.
+The results from @fig-delay-by-line shows that blue network routes (@ttcRoutesSchedule) buses incur the highest average delay.
 This result makes sense as compared to the day shift buses, there are far fewer night shift buses resulting in fewer drivers for various locations the buses may take.
@@
 These findings are what we might expect as there is fewer traffic delays during the night resulting in a more consistent decreasing delay pattern.
-A possible reason for longer delay duration despite fewer traffic could be the fact there there are far fewer blue-network (night shift) buses that run resulting in fewer drivers that require to cover longer distances.
+A possible reason for longer delay duration despite fewer traffic could be the fact there there are far fewer blue netwrok routes (@ttcRoutesSchedule) buses that run resulting in fewer drivers that require to cover longer distances.
 The graph shows that as we get later into the night shift (i.e early morning) we notice the delays dropping as low as roughly 14 minutes.
 
-## Third discussion point
+## Annual frequency of reported delay incidents
+The results from @fig-freq-delay-incidents showcase the total count of instances of delays by various incident reports in the year 2023.
+The above bar graph shows that the most occurring incident was due to operational delays accounting for 2467 total reports.
+The least occurring incident was Rail switches accounting for only 21 total instance reports in the year.
 
-## Weaknesses and next steps
+## Annual delay in minutes by seasons in a year
+The results from @fig-delay-by-season show that on average there wasn't as much variability in delays between the various seasons in a year with summer month accounting for the longest reported average delays.
+Interestingly, the winter month accounted to the least average delays.
+This result is in-line with the existing literature regarding the weather reports in Canada during the winter season as for the first time in history, Toronto reported an average temperature above -5 degrees throughout the month of December and also reported a warmer winter season overall (@canadaWinterArticle).
 
-Weaknesses and next steps should also be included.
+## Conclusion
+By using the findings from @fig-delay-by-line, and @fig-delay-nightshift, we can conclude that there on average the night shift streetcars showcased the longest average delays with periods between 2 AM to 4 AM having the most delays.
+These findings support our understanding that there are far fewer night shift buses that operate on average compared to the number of streetcars that operate during the day's regular working hours.
+@fig-delay-dayshift reports that the day shift streetcars have their reported delays being high during peak hours like 8 AM, 11 AM, 1 PM, 5 PM or 8 PM.
+These results are in-line with the idea that hours between 9-5 are the general working hours in a day and likely can result in high traffic volumes for individuals and commuters.
 
-\newpage
+In addition to studying the variation in delays by the working shift, their hours of day and by streetcar lines, the study also aims to examine the causes of delays and whether there are any seasonal variations that may attribute to potential delays in TTC streetcars.
+The findings from @fig-freq-delay-incidents report that during the year 2023, operation delays were the most reported instances of delays of 2467.
+Furthermore, from @fig-delay-by-season, the study finds that for the year 2023 there weren't as much variability in average delays of streetcars between the seasons.
+This result makes sense as during the winter season due to the harsh weather conditions, there is expected operational delays between streetcar arrivals as seen by the article in 2015 that shows that 25 streetcars were knocked off route (@streetcarOffRoute).
 
-\appendix
+The results from our study highlight important considerations on the types of delay incidents that occur in streetcars, what time of day between various working shifts are there highest reported delays and whether season has any impact on the cause of delays.
+By examining these results, we hope to be able to allocate improved resources to address the various reported incidents for delays.
+To address the issue of delays during peak hours, a potential solution might be to employ additional streetcars that run during these hours and by investing more resources in the night shift streetcars the workload and reduce long delays during late hours like 2-4 AM as these hours are usually dangerous to be out especially alone.
 
-# Appendix {-}
+# Limitations
 
-# Additional data details
+Despite being able to extract several key insights as mentioned in the [discussion] section, our study does have certain limitations that need to be addressed as part of next steps.
+
+One of the key limitation is due to the consequence of our dataset.
+It is important to highlight that the results and inferences made part of this study are for the year 2023 and cannot be generalized to the overall trend in delay patterns that we may observe.
+The results from @fig-delay-by-season is one such example as the result will likely have strong correlation to the observed weather data for the specific year.
+During the year 2023, Toronto reported a unusually warmer time resulting in less frequent and intense snowfall thus not negatively impacting streetcar track or traffic significantly.
+It is important to note that in order to make stronger inferences from our data, we need to extend our dataset to account for past year's of TTC delay data collected from OpenData Toronto.
+
+Another limitation is relating to the quality of the data that we may have.
+Every year TTC reports its Annual Service Plan (@ttc2024Annual) which highlights the track maintenance works and re-routes that may occur during the summer and fall months.
+As part of this, many routes are completely closed down and as such they switch over to alternative modes of transportation like using shuttle buses.
+When collecting data from various years, it is important to make sure that any lack of available data is mentioned in our study as data for these maintenance months may be missing.
+
+# Next Steps {#nextsteps}
+
+In terms of the next steps, there are 2 ways we could extend our study to improve the findings from it.
+
+The first is by addressing the above mentioned limitation of combining data from various years to provide a more richer quality of identified results.
+By combining the data from past years, we are better able to learn of any temporal trends in the data which can lead to interesting results.
+
+The second way to improve our study is by making use of available streetcar delay sentiment data to understand the recurring pain points and emotions that are expressed by commuters.
+This data when combined with existing streetcar delay data can provide richer contextual information about the types of incident delays and the sentiment expressed.
+Another interesting result we can hope to learn is how these sentiments vary between various seasons as during harsher seasons like summer and winter, one can imagine that it wouldn't be most pleasant to experience a long delay in streetcar arrivals.
+
+These findings can help to better connect with the population and provide valuable information on how to better modify or establish new policies that can help to improve the overall commuting experience for the residents of Toronto, the city they call home.
 

### 2024-09-26 scripts/00-simulate_data.R

@@ -49,7 +49,7 @@ data <-
 
 
 #### Write_csv
-write_csv(data, file = "data/raw_data/simulated_streetcar_delay.csv")
+write_csv(data, file = "data/simulated_data/simulated_streetcar_delay.csv")
 
 
 

### 2024-09-26 scripts/02-data_cleaning.R

@@ -14,6 +14,9 @@ set.seed(20136)
 #### Clean data ####
 raw_data <- read_csv("data/raw_data/unedited_ttc_streetcar_delay_2023.csv")
 
+# Report total number of missing values -- referenced in paper.pdf dataset description
+length(which(is.na(raw_data)))
+
 cleaned_data <-
   raw_data |>
   janitor::clean_names() |> 
@@ -21,7 +24,7 @@ cleaned_data <-
            into = c("year", "month"),
            sep = "-") |>
   separate(col = time,
-           into = c("hour", "minute"),
+           into = c("hour"),
            sep = ":") |>
   mutate(season = case_when(
     month %in% c("03", "04", "05") ~ "Spring",  # March, April, May
@@ -31,13 +34,13 @@ cleaned_data <-
   )) |>
   drop_na() |>  # Ensure drop_na() is applied to the cleaned_data
   filter(min_delay != 0) |>
-  filter(min_gap != 0) |>
-  select(year, month, hour, minute, incident, min_delay, line, season)
+  select(year, month, hour, incident, min_delay, line, season)
 
+nrow(cleaned_data)
 # Since the delay time for some dates are over many hours, we will restrict
 # ourselves to at most 2 hour delays in our study
 cleaned_data <- cleaned_data |> filter(min_delay < 121)
-
+nrow(cleaned_data)
 # Check if any column contains NA data or not
 any(is.na(cleaned_data))
 

### 2024-09-26 scripts/03-test_data.R

@@ -13,7 +13,7 @@ library(tidyverse)
 set.seed(20136)
 
 #### Test data ####
-data <- read_csv("data/raw_data/simulated_streetcar_delay.csv")
+data <- read_csv("data/simulated_data/simulated_streetcar_delay.csv")
 
 # Test for negative numbers in delay_mins column
 data$delay_mins |> min() <= 0

### 2024-09-26 paper/paper.qmd
Update references, clean up graphs and fix citation issues

@@
 bibliography: references.bib
-link-citations: true
 ---
@@
 library(gt)
+library(GGally)
 
@@
 In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced.
-We make use R (@citeR) and the dataset made available by OpenData Toronto (@opendatatoronto) for the year 2023.
+In this paper, we make use R (@citeR) and its supported packages like tidyverse  (@tidyverse) and dplyr (@dplyr).
+For our dataset, we use the provided data for the year 2023 by OpenData Toronto (@opendatatoronto).
 The reported graphs in our study is generated using ggplot (@ggplot2) and the tables are generated using the gt (@gt) library.
@@
 The remainder of this paper is structured as follows.
-We first give a brief overview of the dataset under [datasetdesc] section.
-Then, we present some graphs on the results extracted from our collected data found under [results] section.
-Lastly, we conclude with a discussion of our results found under [discussion] section, highlight the limitations in our conducted study found under [limitations] section and propose possible future next steps for this project under [nextsteps] section.
+We first give a brief overview of the dataset under [dataset description](#datasetdesc) section.
+Then, we present some graphs on the results extracted from our collected data found under [results](#results).
+We then proceed to a discussion of our results found under [discussion](#discussion) section and highlight the limitations in our conducted study found under [limitations](#limitations) section and end our study with some proposed [next steps](#nextsteps).
 
@@
 Therefore, no transformations are necessary to populate missing entries in our dataset.
-Lastly, as highlighed in the [assumption] section, we filter our dataset to only include delays of upto 120 minutes.
+Lastly, as highlighed in the [assumption](#assumption) section, we filter our dataset to only include delays of upto 120 minutes.
 
@@
 ggplot(grp_data, aes(x = factor(line), y = average_delay)) +
-  geom_bar(stat = "identity", width = 0.7, fill = 'lightblue') +
-  geom_text(aes(label = round(average_delay, 2)), 
+  geom_bar(stat = "identity", width = 0.7, fill = 'dodgerblue1', alpha = 0.75) +
+  geom_text(aes(label = paste0(round(average_delay, 2))),
             vjust = -0.5,  # Position the text above the bar
-            size = 2) +  # Adjust the size of the text
+            size = 1.75) +  # Adjust the size of the text
   ggtitle("Average delay in minutes by streetcar line") +
@@
 
-ggplot(avg_delay_day_shift, aes(x = hour, y = average_delay)) +
-  geom_line(color = "red", size = 1) +
-  geom_point(color = "blue", size = 2) +
-  geom_text(aes(label = round(average_delay, 2)), 
-            vjust = -1, 
-            size = 2) + 
+ggplot(avg_delay_day_shift, aes(x = as.numeric(hour), y = average_delay)) +
+  geom_line(color = "dodgerblue1", size = 1, alpha = 0.5) +
+  geom_point(color = "dodgerblue4", size = 2) +
+  geom_text(aes(label = round(average_delay, 1)), 
+            nudge_y = 0.25, 
+            size = 1.75) + 
   ggtitle("Average delay in minutes for morning shift") +
@@
 # Plot the graph with the reordered x-axis
-ggplot(avg_delay_night_shift, aes(x = hour, y = average_delay)) +
-  geom_line(color = "blue", size = 1) +
-  geom_point(color = "red", size = 2) +
-  geom_text(aes(label = round(average_delay, 2)), 
-            vjust = -1, 
-            size = 2) + 
+ggplot(avg_delay_night_shift, aes(x = as.numeric(hour), y = average_delay)) +
+  geom_line(color = "red1", size = 1, alpha = 0.5) +
+  geom_point(color = "red3", size = 2) +
+  geom_text(aes(label = round(average_delay, 1)), 
+            vjust = 2.5, 
+            size = 1.75) + 
   ggtitle("Average delay in minutes for night shift") +
@@
 
+# Calculate percentages
+freq_incidence$percentage <- freq_incidence$frequency / sum(freq_incidence$frequency) * 100
+
+# Plot with percentage labels
 ggplot(freq_incidence, aes(x = incident, y = frequency)) +
-  geom_bar(stat = "identity", fill = "lightblue") +
-  geom_text(aes(label = frequency), 
-            vjust = -0.5,  # Position the text above the bar
-            size = 2) +  # Adjust the size of the text
-  ggtitle("Frequency of each delay incident in the year 2023") +
+  geom_bar(stat = "identity", fill = "dodgerblue1", alpha = 0.75) +
+  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
+            vjust = -0.5,
+            size = 2) +
+  ggtitle("Frequency and percentage of incident types in 2023") +
   xlab("Delay Incident Type") +
@@
     plot.title = element_text(hjust = 0.5),  # Center title
-    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)  # Rotate x-axis labels for clarity
   )
@@
 @fig-freq-delay-incidents highlights the total annual delays incurred based on incidents reported by TTC for the reason.
-We can see for the year of 2023, operation delays were the most frequently occurring including a total of 2463 instances and rail switches were the least occuring with reported instances being only about 20.
+We can see for the year of 2023, 23% of delay incidents were due to operation delays amounting to a total of 2463 instances and rail switches were the least occurring with reported instances being only about 0.2% of all delays.
 
@@
 #| label: fig-delay-by-season
-#| fig-cap: Average delay in minutes by seasons in a yea
+#| fig-cap: Average delay in minutes by seasons in a year
 #| echo: false
@@
 ggplot(delay_by_season, aes(x = season, y = average_delay)) +
-  geom_bar(stat = "identity", fill = "lightblue") +
+  geom_bar(stat = "identity", fill = "dodgerblue1", alpha = 0.75) +
+  geom_text(aes(label = paste0(round(average_delay, 2))), 
+            vjust = -0.5,
+            size = 2) +
   ggtitle("Average delay in minutes by seasons in a year") +
@@
 A possible reason for longer delay duration despite fewer traffic could be the fact there there are far fewer blue netwrok routes (@ttcRoutesSchedule) buses that run resulting in fewer drivers that require to cover longer distances.
-The graph shows that as we get later into the night shift (i.e early morning) we notice the delays dropping as low as roughly 14 minutes.
+The graph shows that as we get later into the night shift, we notice the delays dropping as low as roughly 14 minutes.
 
@@
 Interestingly, the winter month accounted to the least average delays.
-This result is in-line with the existing literature regarding the weather reports in Canada during the winter season as for the first time in history, Toronto reported an average temperature above -5 degrees throughout the month of December and also reported a warmer winter season overall (@canadaWinterArticle).
+This result is in-line with the existing literature regarding the weather reports in Canada during the winter season as for the first time in history, Toronto reported an average temperature above -5 degrees Celsius throughout the month of December and also reported a warmer winter season overall (@canadaWinterArticle).
 
@@
 Furthermore, from @fig-delay-by-season, the study finds that for the year 2023 there weren't as much variability in average delays of streetcars between the seasons.
-This result makes sense as during the winter season due to the harsh weather conditions, there is expected operational delays between streetcar arrivals as seen by the article in 2015 that shows that 25 streetcars were knocked off route (@streetcarOffRoute).
+This result makes sense as during the winter season due to the harsh weather conditions, there is expected operational delays between streetcar arrivals as seen by the article in 2015 that shows that 25 streetcars were knocked off route (@torontosunColdWeather).
 
@@
 
-# Limitations
+# Limitations {#limitations}
 
-Despite being able to extract several key insights as mentioned in the [discussion] section, our study does have certain limitations that need to be addressed as part of next steps.
+Despite being able to make various conclusions as mentioned in the [discussion](#discussion) section, our study does have certain limitations that need to be addressed as part of next steps.
 

### 2024-09-26 paper/paper.qmd
Updated llm usage.txt and removed unused files/folders

@@
   group_by(line) |>
-  summarise(average_delay = mean(min_delay), .groups = 'drop')
+  summarise(average_delay = mean(min_delay))
 
@@
   group_by(hour) |>
-  summarise(average_delay = mean(min_delay), .groups = 'drop') |>
+  summarise(average_delay = mean(min_delay)) |>
   filter(hour >= "08" | hour <= "01")
@@
   filter(hour >= "02" & hour <= "07") |>
-  summarise(average_delay = mean(min_delay), .groups = 'drop')
+  summarise(average_delay = mean(min_delay))
 
@@
   group_by(incident) |> 
-  summarise(frequency = n(), .groups = 'drop')
+  summarise(frequency = n())
 
@@
   group_by(season) |>
-  summarise(average_delay = mean(min_delay), .groups = 'drop')
+  summarise(average_delay = mean(min_delay))
 

### 2024-09-27 paper/paper.qmd
Added citation for professor's starter code and updated sketches
Update data cleaning to convert numbers to numeric type, add details about measurement and cleaned_data update

@@
 library(gt)
-library(GGally)
 
@@
 In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced.
-In this paper, we make use R (@citeR) and its supported packages like tidyverse  (@tidyverse) and dplyr (@dplyr).
+The starter code for this paper was provided by Professor Rohan (@citeRohan).
+We make use R (@citeR) and its supported packages like tidyverse  (@tidyverse) and dplyr (@dplyr).
 For our dataset, we use the provided data for the year 2023 by OpenData Toronto (@opendatatoronto).
@@
 
-The following table describes the data types for the selected features in our study from the cleaned dataset.
+The data is collected and maintained by TTC and is updated monthly as reported by the website.
+Since the data is stored as a CSV, several numerical fields like the streetcar line or delay in minutes are saved as characters instead of numbers.
+Fields like date are saved as appropriate format of datetime type.
+
+The dataset was last updated September 24, 2024 ensuring consistent updates to the collected data.
+OpenData Toronto also provides functionality and sufficient instructions for developers to easily query their data using scripts in Python, R and Node.js instead of having to manually download them.
+
+The website doesn't provided information on how the data was sourced but we can assume that the data is recorded by taking the difference in the reported arrival time via the TTC app and the actual time the streetcar arrives at various stops.
+The collected data represents common types like characters, numeric and datetime.
+Therefore, for the purposes of this study, the procedure and type of the data reported is appropriate for our use and we do not require any extra information to make any assumptions about the data types.
+
+The following table describes the data types for the selected features in our study from the cleaned dataset after the necessary transformations.
 ```{r}
@@
 
-# Filter the data for times between 08:00 and 01:00 i.e day shift
+# Filter the data for times between 08:00 and 00:00 i.e day shift
 avg_delay_day_shift <- data |> 
@@
   summarise(average_delay = mean(min_delay)) |>
-  filter(hour >= "08" | hour <= "01")
+  filter(hour >= 8 & hour <= 23)
 
+# Plot the graph
 ggplot(avg_delay_day_shift, aes(x = as.numeric(hour), y = average_delay)) +
@@
   ylab("Average Delay (Minutes)") +
+  scale_x_continuous(breaks = seq(8, 23)) +
   theme_minimal() +
@@
 @fig-delay-dayshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the day.
-We observe that there is high variability in the delays throughout the day with the minimum delay being about 13.78 minutes and the maximum being roughly 20 minutes.
+We observe that there is high variability in the delays throughout the day with the minimum delay being about 12.2 minutes at 8 AM and the maximum being roughly 17 minutes at 11 PM.
 
@@
 
-# Filter the data for times between 02:00 and 07:00 i.e night shift
+# Filter the data for times between 00:00 and 07:00 i.e night shift
 avg_delay_night_shift <- data |> 
   group_by(hour) |>
-  filter(hour >= "02" & hour <= "07") |>
+  filter(hour >= 0 & hour <= 7) |>
   summarise(average_delay = mean(min_delay))
 
-# Plot the graph with the reordered x-axis
-ggplot(avg_delay_night_shift, aes(x = as.numeric(hour), y = average_delay)) +
+# Plot the graph
+ggplot(avg_delay_night_shift, aes(x = hour, y = average_delay)) +
   geom_line(color = "red1", size = 1, alpha = 0.5) +
@@
   ylab("Average Delay (Minutes)") +
+  scale_x_continuous(breaks = seq(0, 7)) + 
   theme_minimal() +
@@
 @fig-delay-nightshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the night.
-We observe that there is decreasing trend in the delays throughout the night after 3 AM with the minimum delay being about 14 minutes and the maximum roughly 27 minutes.
+We observe that there is maximum delay at 3 AM for roughly 25 minutes and the minimum delay being about 11.4 minutes at 7 AM.
 

### 2024-09-27 scripts/02-data_cleaning.R

@@ -32,15 +32,23 @@ cleaned_data <-
     month %in% c("09", "10", "11") ~ "Fall",     # September, October, November
     month %in% c("12", "01", "02") ~ "Winter"    # December, January, February
   )) |>
-  drop_na() |>  # Ensure drop_na() is applied to the cleaned_data
+  drop_na() |>
   filter(min_delay != 0) |>
   select(year, month, hour, incident, min_delay, line, season)
 
+# Ensure that the numbers are converted to numeric from character (if that's the case)
+cleaned_data$hour <- as.numeric(cleaned_data$hour)
+cleaned_data$month <- as.numeric(cleaned_data$month)
+cleaned_data$year <- as.numeric(cleaned_data$year)
+cleaned_data$min_delay <- as.numeric(cleaned_data$min_delay)
+cleaned_data$line <- as.numeric(cleaned_data$line)
+
 nrow(cleaned_data)
 # Since the delay time for some dates are over many hours, we will restrict
-# ourselves to at most 2 hour delays in our study
+# ourselves to at most 2 hour delays in our study -- discussed in assumption section of the paper
 cleaned_data <- cleaned_data |> filter(min_delay < 121)
 nrow(cleaned_data)
+
 # Check if any column contains NA data or not
 any(is.na(cleaned_data))
 

### 2024-09-27 paper/paper.qmd
Update ReadMe, scripts for test and simulation and llm usage. Recompile paper and qmd document

@@
 
+\newpage
 ## Data preprocessing and cleaning
-The raw data is undergone a series of data cleaning steps to ensure adequate usability for the study.
-We first use the Janitor (@janitor) package to clean the data columns removing any spaces between them and using a consistent naming stype for them.
-Next, we proceed to split the date field into year and month fields and split the time field extracting the hour of day from it.
-Following this, we create a new column to denote the season based on the month of the year from the data.
-This helps during the analysis stage when we need to make inferences about the seasonal variation in average delays in streetcars.
-
-We then proceed to drop the rows with missing values or NA.
-Out of the total 14413 rows of data, 2399 of them contain NA values in total which doesn't heavily impact our data.
-Therefore, no transformations are necessary to populate missing entries in our dataset.
-Lastly, as highlighed in the [assumption](#assumption) section, we filter our dataset to only include delays of upto 120 minutes.
-
-As a result of the above described data cleaning process, we select our desired features from the dataset. The following table shows the first five rows of our data and the selected features.
 
@@
 #| echo: false
+#| label: tbl-dataset-preview
+#| tbl-cap: "First first five rows in the cleaned dataset with selected features"
 
@@
   )
-cat("Table 1: First first five rows in the cleaned dataset with selected features")
 
@@
 
+The raw data is undergone a series of data cleaning steps to ensure adequate usability for the study.
+We first use the Janitor (@janitor) package to clean the data columns removing any spaces between them and using a consistent naming stype for them.
+Next, we proceed to split the date field into year and month fields and split the time field extracting the hour of day from it.
+Following this, we create a new column to denote the season based on the month of the year from the data.
+This helps during the analysis stage when we need to make inferences about the seasonal variation in average delays in streetcars.
+
+We then proceed to drop the rows with missing values or NA.
+Out of the total 14413 rows of data, 2399 of them contain NA values in total which doesn't heavily impact our data.
+Therefore, no transformations are necessary to populate missing entries in our dataset.
+Lastly, as highlighed in the [assumption](#assumption) section, we filter our dataset to only include delays of upto 120 minutes.
+
+As a result of the above described data cleaning process, we select our desired features from the dataset. @tbl-dataset-preview shows the first five rows of our data and the selected features.
+
+\newpage
 ## Measurement of units for selected features
 
-The data is collected and maintained by TTC and is updated monthly as reported by the website.
-Since the data is stored as a CSV, several numerical fields like the streetcar line or delay in minutes are saved as characters instead of numbers.
-Fields like date are saved as appropriate format of datetime type.
-
-The dataset was last updated September 24, 2024 ensuring consistent updates to the collected data.
-OpenData Toronto also provides functionality and sufficient instructions for developers to easily query their data using scripts in Python, R and Node.js instead of having to manually download them.
-
-The website doesn't provided information on how the data was sourced but we can assume that the data is recorded by taking the difference in the reported arrival time via the TTC app and the actual time the streetcar arrives at various stops.
-The collected data represents common types like characters, numeric and datetime.
-Therefore, for the purposes of this study, the procedure and type of the data reported is appropriate for our use and we do not require any extra information to make any assumptions about the data types.
-
-The following table describes the data types for the selected features in our study from the cleaned dataset after the necessary transformations.
 ```{r}
@@
 #| echo: false
+#| label: tbl-selected-features
+#| tbl-cap: "Datatypes for the selected features from the cleaned dataset"
 
@@
   )
-  cat("Table 2: Datatypes for the selected features from the cleaned dataset.")
 ```
+
+The data is collected and maintained by TTC and is updated monthly as reported by the website.
+Since the data is stored as a CSV, several numerical fields like the streetcar line or delay in minutes are saved as characters instead of numbers.
+Fields like date are saved as appropriate format of datetime type.
+
+The dataset was last updated September 24, 2024 ensuring consistent updates to the collected data.
+OpenData Toronto also provides functionality and sufficient instructions for developers to easily query their data using scripts in Python, R and Node.js instead of having to manually download them.
+
+The website doesn't provided information on how the data was sourced but we can assume that the data is recorded by taking the difference in the reported arrival time via the TTC app and the actual time the streetcar arrives at various stops.
+The collected data represents common types like characters, numeric and datetime.
+Therefore, for the purposes of this study, the procedure and type of the data reported is appropriate for our use and we do not require any extra information to make any assumptions about the data types.
+
+@tbl-selected-features describes the data types for the selected features in our study from the cleaned dataset after the necessary transformations.
 

### 2024-09-27 scripts/00-simulate_data.R

@@ -13,38 +13,35 @@ library(tidyverse)
 set.seed(20136)
 
 #### Simulate data ####
-# Define the start and end date
-start_date <- as.Date("2023-01-01")
-end_date <- as.Date("2023-12-31")
-
 # Set the number of random dates you want to generate
-number_of_dates <- 100
-from_stations <- c("Finch", "Sheppard Yongue", "Lawrence Station", "Bloor Station")
-to_stations <- c("St George Station", "Union Station", "Queen Station", "Dundas West", "Museum")
+num_rows <- 100
 
 data <-
   tibble(
-    dates = as.Date(
-      runif(
-        n = number_of_dates,
-        min = as.numeric(start_date),
-        max = as.numeric(end_date)
+    year = 2023,
+    month = as.integer(runif(
+      n = num_rows,
+      min = 1,
+      max = 12
+    )),
+    hour = as.integer(runif(
+      n = num_rows,
+      min = 0,
+      max = 23
+    )),
+    delay_mins = as.integer(runif(
+      n = num_rows,
+      min = 1,
+      max = 120
+    )),
+    streetcar_line = sample(
+      c(504, 505, 506, 511, 304, 305, 311, 9000),
+      size = num_rows, replace = TRUE
       ),
-    ),
-    delay_mins = rpois(n = number_of_dates, lambda = 20),
-    day = weekdays(dates),
     reason = sample(
       c("Traffic", "Emergency", "Weather", "Route Change"), 
-      size = number_of_dates, replace = TRUE
-      ),
-    from_station = sample(from_stations, 
-                          size = number_of_dates, 
-                          replace = TRUE
-    ),
-    to_station = sample(to_stations, 
-                        size = number_of_dates, 
-                        replace = TRUE
-    ),
+      size = num_rows, replace = TRUE
+      )
   )
 
 

### 2024-09-27 scripts/02-data_cleaning.R

@@ -14,9 +14,6 @@ set.seed(20136)
 #### Clean data ####
 raw_data <- read_csv("data/raw_data/unedited_ttc_streetcar_delay_2023.csv")
 
-# Report total number of missing values -- referenced in paper.pdf dataset description
-length(which(is.na(raw_data)))
-
 cleaned_data <-
   raw_data |>
   janitor::clean_names() |> 
@@ -43,14 +40,9 @@ cleaned_data$year <- as.numeric(cleaned_data$year)
 cleaned_data$min_delay <- as.numeric(cleaned_data$min_delay)
 cleaned_data$line <- as.numeric(cleaned_data$line)
 
-nrow(cleaned_data)
 # Since the delay time for some dates are over many hours, we will restrict
 # ourselves to at most 2 hour delays in our study -- discussed in assumption section of the paper
 cleaned_data <- cleaned_data |> filter(min_delay < 121)
-nrow(cleaned_data)
-
-# Check if any column contains NA data or not
-any(is.na(cleaned_data))
 
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/cleaned_ttc_streetcar_delay_2023.csv")

### 2024-09-27 scripts/03-test_data.R

@@ -12,18 +12,39 @@
 library(tidyverse)
 set.seed(20136)
 
-#### Test data ####
-data <- read_csv("data/simulated_data/simulated_streetcar_delay.csv")
+#### Test data for cleaned dataset ####
+data <- read_csv("data/analysis_data/cleaned_ttc_streetcar_delay_2023.csv")
 
-# Test for negative numbers in delay_mins column
-data$delay_mins |> min() <= 0
-# Test for NAs in delay_mins column
-all(is.na(data$delay_mins))
-# Test for NAs in date column
-all(is.na(data$dates))
-# Test for NAs in reason column
-all(is.na(data$reason))
-# Test for NAs in to_station column
-all(is.na(data$to_station))
-# Test for NAs in from_station column
-all(is.na(data$from_station))
+# Test for whether any column has missing/NA values
+any(is.na(data))
+
+# Check if all the datatypes are of valid type
+if (!is.character(data$season)) stop("seasons should be character")
+if (!is.numeric(data$line)) stop("line should be numeric")
+if (!is.numeric(data$min_delay)) stop("min_delay should be numeric")
+if (!is.numeric(data$year)) stop("year should be numeric")
+if (!is.numeric(data$month)) stop("month should be numeric")
+if (!is.numeric(data$hour)) stop("hour should be numeric")
+if (!is.character(data$incident)) stop("incident should be character")
+
+# Check if months are within valid range
+if (any(data$month < 1 | data$month > 12)) stop("Month values must be between 1 and 12")
+
+# Check if year is 2023
+if (any(data$year != 2023)) stop("Year should be 2023")
+
+# Check hour values in military style
+if (any(data$hour < 0 | data$hour > 23)) stop("Hour values must be between 0 and 23")
+
+# Check min_delay is at least 1 minute
+if (any(data$min_delay < 0)) stop("Delays must be non-negative")
+
+# Check min_delay is not greater than 120 mins as per the assumption specified
+if (any(data$min_delay > 120)) stop("Delays must be less than 120 minutes")
+
+# Check seasons
+valid_seasons <- c("Winter", "Spring", "Summer", "Fall")
+if (any(!data$season %in% valid_seasons)) stop("Invalid season values detected")
+
+# Check incident descriptions
+if (any(nchar(data$incident) == 0)) stop("Incident descriptions must be non-empty")

### 2024-09-27 paper/paper.qmd
Improved wordings in report and readme

@@
 date-format: long
-abstract: "This paper aims to study the causes of delays in Toronto Transit Commissions (TTC) streetcar service, the variation in delays by the hours in working shifts and whether there are any seasonal relationships to the observed delays for the year 2023.
-By using the data provided by OpenData Toronto, we are able to understand the observed relationships between the several fields of interests like minimum delay, hour, season, streetcar line and incident type.
-Our study finds that there is higher reported delayed times during night shift and higher delays reported during peak hours that fall during the working 9AM - 5PM time period.
-Interestingly we did not observe any variation in delays during the various seasons and understand that to make stronger inferences between the temporal relationship, we need to examine longer historic data."
+abstract: "This paper aims to study the causes of delays in Toronto Transit Commissions (TTC) streetcar service by examining the relationship between various factors like hour of day, streetcar lines and seasons.
+Our study found that the night shift had the higest delays by average of 25 minutes with high variability in delays observed during the hours of 9 AM to 5 PM.
+Interestingly, we did not observe any seasonal variations in the overall delay in streetcar and one possible reason for this could be due to our limited dataset.
+We beleive the results from this paper can help in better designing policies and transit schedules to improve the overall commuting experience."
 format: 
@@
 
-These delays make it especially difficult for individuals that commute on a daily basis.
-Furthermore, one can image the inconveniences caused due to extended delays by streetcars during weather conditions like the winter months.
+The delays via streetcars make it especially difficult for commuters on a daily basis.
+One can image the inconveniences by these delays during the harsh winter weather conditions and the struggles commuters face.
+In 2019, streetcars averaged delays of about 13,790 minutes each month and by 2023, this number increased by 33% (@streetcarDelay2023).
+As a result, commuters are in need for improved policies and better transit schedules to address the difficulties they face everyday.
+Our study found that the night shift had the higest delays by average of 25 minutes with high variability in delays observed during the hours of 9 AM – 5 PM.
+Interestingly, we did not observe any seasonal variations in the overall delay in streetcar and one possible reason for this could be due to our limited dataset.
+
 In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced.
@@
 We make use R (@citeR) and its supported packages like tidyverse  (@tidyverse) and dplyr (@dplyr).
-For our dataset, we use the provided data for the year 2023 by OpenData Toronto (@opendatatoronto).
-The reported graphs in our study is generated using ggplot (@ggplot2) and the tables are generated using the gt (@gt) library.
-Code cleanup is performed using the Janitor (@janitor) package.
+For our dataset, we use the TTC streetcar delay data for the year 2023 by OpenData Toronto (@opendatatoronto).
+The reported graphs in our study are generated using ggplot (@ggplot2) and the tables are generated using the gt (@gt) package.
+The process to perform data cleaning was done using the Janitor (@janitor) package.
 
 The remainder of this paper is structured as follows.
-We first give a brief overview of the dataset under [dataset description](#datasetdesc) section.
-Then, we present some graphs on the results extracted from our collected data found under [results](#results).
-We then proceed to a discussion of our results found under [discussion](#discussion) section and highlight the limitations in our conducted study found under [limitations](#limitations) section and end our study with some proposed [next steps](#nextsteps).
+We first give a brief overview of the dataset under [dataset description](#datasetdesc) which talks about the various features, the assumptions we make and the measurements done to gather the data.
+We present some graphs on the results extracted from our collected data found under [results](#results).
+A discussion of our results can be found under [discussion](#discussion) section with the relevant limitations in our study found under [limitations](#limitations) section.
+We end our study with some proposed [next steps](#nextsteps) for future version of this paper.
 
@@
 The following data was made available from OpenData Toronto (@opendatatoronto) for the year 2023.
-The data was extracted from the server and saved locally using in a raw data folder.
+The data was extracted from the server and saved locally using in a raw data folder using the instructions provided by the documentation.
+We note that this was the only available dataset with frequent updates and data dating back to 2014.
 Some of the features part of this dataset are date, time, incident, min delay, bound and location.
@@
 These values can heavily skew the results that we may observe.
-Therefore, one key assumption we make in our study is that we limit the maximum delay time to upto 2 hours or 120 minutes.
+Therefore, one key assumption we make in our study is that we are only interested in at most a delay by 2 hours in streetcars.
+Therefore, during the data cleaning step, we will limit the rows of data to 120 minutes.
 As a result of this filtering, we only lose about 130 data and still retain about 10723 data entries to use in our study.
@@
 The raw data is undergone a series of data cleaning steps to ensure adequate usability for the study.
-We first use the Janitor (@janitor) package to clean the data columns removing any spaces between them and using a consistent naming stype for them.
-Next, we proceed to split the date field into year and month fields and split the time field extracting the hour of day from it.
+We first use the Janitor (@janitor) package to clean the data columns removing any spaces between them and using a consistent variable name convention.
+Next, we proceed to split the date field into year and month fields and do the same with the time field extracting the hour of day from it.
 Following this, we create a new column to denote the season based on the month of the year from the data.
-This helps during the analysis stage when we need to make inferences about the seasonal variation in average delays in streetcars.
+This is used for studying any seasonal variation in delays that may exist in our data.
 
-We then proceed to drop the rows with missing values or NA.
-Out of the total 14413 rows of data, 2399 of them contain NA values in total which doesn't heavily impact our data.
-Therefore, no transformations are necessary to populate missing entries in our dataset.
-Lastly, as highlighed in the [assumption](#assumption) section, we filter our dataset to only include delays of upto 120 minutes.
+After selecting the appropriate columns in our dataset, we proceed to drop the rows with missing values in them.
+Out of the total 14413 rows of data, 2399 of them contain missing values.
+As a consequence of dropping these rows, we do not lose any valuable information in our dataset.
+Therefore, no transformations are necessary to populate any missing entries.
+Lastly, as highlighted in the [assumption](#assumption) section, we filter our data to only include delays up to 120 minutes.
 
-As a result of the above described data cleaning process, we select our desired features from the dataset. @tbl-dataset-preview shows the first five rows of our data and the selected features.
+@tbl-dataset-preview shows the first five rows of our data and the selected features.
 
@@
 The dataset was last updated September 24, 2024 ensuring consistent updates to the collected data.
-OpenData Toronto also provides functionality and sufficient instructions for developers to easily query their data using scripts in Python, R and Node.js instead of having to manually download them.
+OpenData Toronto also provides functionality and sufficient instructions for developers to easily query their data using scripts in Python (@python), R (@citeR) and Node.js (@nodeJS) instead of having to manually download them.
 
@@
 The collected data represents common types like characters, numeric and datetime.
-Therefore, for the purposes of this study, the procedure and type of the data reported is appropriate for our use and we do not require any extra information to make any assumptions about the data types.
+Therefore, for the purposes of this study, the procedures used to collect the data and their types are appropriate for our use and we do not require any extra information to make any assumptions them.
 
-@tbl-selected-features describes the data types for the selected features in our study from the cleaned dataset after the necessary transformations.
+@tbl-selected-features describes the data types for the selected features in our study from the cleaned dataset.
 
@@
 #| label: fig-delay-by-line
-#| fig-cap: Average annual delay in minutes for each streetcar line
+#| fig-cap: Average annual delay in minutes for each streetcar line in 2023
 
@@
 @fig-delay-by-line highlights the annual average delay in minutes for every streetcar line for the year 2023.
-The 300-series lines also called the Blue Network Routes (@ttcRoutesSchedule) are the nighttime streetcar lines that start at 1:30 AM.
-These buses average the most delays reaching at most 30 minutes for line 312.
+The 300-series lines also called the Blue Network Routes (@ttcRoutesSchedule) are the night time streetcar lines that start at 12:00 AM.
+These average the most delays reaching at most 30 minutes for line 312.
 
-The 500-series lines are the daytime lines that run from 8 AM. These buses average between 6 to 23.6 minutes with the longest delay experienced by line 507.
+The 500-series lines are the daytime lines that run from 8 AM. These streetcar average between 6 to 23.6 minutes with the longest delay experienced by line 507.
 
@@
 #| label: fig-delay-dayshift
-#| fig-cap: Average delay in minutes for dayshift
+#| fig-cap: Average delay in minutes for dayshift in 2023
 #| echo: false
@@
 #| label: fig-delay-nightshift
-#| fig-cap: Average delay in minutes for nightshift
+#| fig-cap: Average delay in minutes for nightshift in 2023
 #| echo: false
@@
 ```
-@fig-freq-delay-incidents highlights the total annual delays incurred based on incidents reported by TTC for the reason.
-We can see for the year of 2023, 23% of delay incidents were due to operation delays amounting to a total of 2463 instances and rail switches were the least occurring with reported instances being only about 0.2% of all delays.
+@fig-freq-delay-incidents highlights the total annual delays incurred based on incidents reported by TTC for the cause of said delay.
+We can that 23% of delay incidents were due to operational reasons amounting to a total of 2463 instances and rail switches were the least occurring with reported instances being only about 0.2% of all delays.
 
@@
 #| label: fig-delay-by-season
-#| fig-cap: Average delay in minutes by seasons in a year
+#| fig-cap: Average delay in minutes by seasons in 2023
 #| echo: false
@@
 
-## Delays by bus lines
-The results from @fig-delay-by-line shows that blue network routes (@ttcRoutesSchedule) buses incur the highest average delay.
-This result makes sense as compared to the day shift buses, there are far fewer night shift buses resulting in fewer drivers for various locations the buses may take.
-Additionally, we also note that the 900-series streetcar line that go to the Airport incur the longest average delay of almost 54 minutes.
+## How do delays vary by streetcars lines?
+The results from @fig-delay-by-line shows that blue network routes (@ttcRoutesSchedule) streetcar incur the highest average delay of about 25 minutes.
+This result makes sense as compared to the day shift streetcar, there are far fewer night shift streetcar resulting in fewer drivers for various locations the streetcar may take.
 
-## Average delay for morning shift
+## Are night time streetcars busier than the day?
 The results from @fig-delay-dayshift showcases the variability in average delays by the hour during the morning shift working hours observed between 8 AM to 6 PM.
-We observe that the results are in-line with the expected behavior of longer average delays of around 18 minutes during peak hours like 8 AM or 5 PM which is usually when a large working population leave and return from work.
+We observe that the results are in-line with the expected behavior of longer average delays of around 18 minutes during peak hours like 8 AM or 5 PM which is usually when a large working population leave for and return from work.
 This can result in higher traffic hours on average causing an expected delay in streetcar arrivals.
 Additionally, we notice that throughout the day there are several points of interest for peak delays as seen at 11 AM or 1 PM incurring an average delay of 19 minutes.
-This observation could indicate the need for travel to various locations like grocery stores, perhaps to a friend's house or classes for university students.
-During these hours, there is heavy traffic and these suggest a potential reason for delays.
+This observation could indicate the need for travel to various locations like grocery stores, perhaps to a friend's house or university students commuting for their class.
 
-## Average delay for night shift
-The results from @fig-delay-nightshift shows a decreasing trend in the average delays by the hour during the night shift working hours observed between 2 AM to 7 AM.
+The results from @fig-delay-nightshift show the average delays by the hour during the night shift working hours observed between 12:00 AM to 7 AM.
 These findings are what we might expect as there is fewer traffic delays during the night resulting in a more consistent decreasing delay pattern.
-A possible reason for longer delay duration despite fewer traffic could be the fact there there are far fewer blue netwrok routes (@ttcRoutesSchedule) buses that run resulting in fewer drivers that require to cover longer distances.
-The graph shows that as we get later into the night shift, we notice the delays dropping as low as roughly 14 minutes.
 
-## Annual frequency of reported delay incidents
+Despite daytime being busier than nighttime, why do streetcars at night report the highest reported delays?
+A possible reason for longer delays despite fewer traffic could be the fact there there are far fewer blue network routes (@ttcRoutesSchedule) streetcar that run resulting in fewer drivers that require to cover longer distances.
+The graph shows that as we get later into the night shift, we notice the delays dropping as low as 11 minutes.
+
+## What causes the delays in streetcars?
 The results from @fig-freq-delay-incidents showcase the total count of instances of delays by various incident reports in the year 2023.
-The above bar graph shows that the most occurring incident was due to operational delays accounting for 2467 total reports.
-The least occurring incident was Rail switches accounting for only 21 total instance reports in the year.
+This graph shows that the most occurring incident was due to operational delays accounting for 23% of total reports with rail switches being the least amounting to just 0.2 %.
 
-## Annual delay in minutes by seasons in a year
+## Is it true that winters making commuting especially challenging?
 The results from @fig-delay-by-season show that on average there wasn't as much variability in delays between the various seasons in a year with summer month accounting for the longest reported average delays.
-Interestingly, the winter month accounted to the least average delays.
-This result is in-line with the existing literature regarding the weather reports in Canada during the winter season as for the first time in history, Toronto reported an average temperature above -5 degrees Celsius throughout the month of December and also reported a warmer winter season overall (@canadaWinterArticle).
+We find it interesting that winter accounted to the least average delays as in 2015, 25 streetcars were knocked off-route due to harsh weather conditions (@torontosunColdWeather).
+One reason for why this wasn't the case in 2023 is due to the fact that Toronto reported an average temperature above -5 degrees Celsius throughout the month of December and also reported a warmer winter season overall (@canadaWinterArticle).
 
 ## Conclusion
-By using the findings from @fig-delay-by-line, and @fig-delay-nightshift, we can conclude that there on average the night shift streetcars showcased the longest average delays with periods between 2 AM to 4 AM having the most delays.
-These findings support our understanding that there are far fewer night shift buses that operate on average compared to the number of streetcars that operate during the day's regular working hours.
+By utilizing the results from @fig-delay-by-line, and @fig-delay-nightshift, we can conclude that on average the night shift streetcars showcased the longest average delays with periods between 2 AM to 4 AM.
+These findings make sense as there are far fewer night shift streetcar that operate on average compared to the number of streetcars that operate during the day.
 @fig-delay-dayshift reports that the day shift streetcars have their reported delays being high during peak hours like 8 AM, 11 AM, 1 PM, 5 PM or 8 PM.
-These results are in-line with the idea that hours between 9-5 are the general working hours in a day and likely can result in high traffic volumes for individuals and commuters.
+These results are in-line with the idea that hours between 9 AM to 5 AM are the general working hours in a day and likely can result in high traffic volumes for commuters.
 
@@
 The results from our study highlight important considerations on the types of delay incidents that occur in streetcars, what time of day between various working shifts are there highest reported delays and whether season has any impact on the cause of delays.
-By examining these results, we hope to be able to allocate improved resources to address the various reported incidents for delays.
-To address the issue of delays during peak hours, a potential solution might be to employ additional streetcars that run during these hours and by investing more resources in the night shift streetcars the workload and reduce long delays during late hours like 2-4 AM as these hours are usually dangerous to be out especially alone.
+By examining these results, we hope to be able to allocate resources to address the various reported incidents for delays.
+To address the issue of delays during peak hours, a potential solution might be to employ additional streetcars that run during these hours and by investing more resources in the night shift to reduce the long delays during late hours like 2 to 4 AM as these hours are usually dangerous to be out especially alone.
 
@@
 
-In terms of the next steps, there are 2 ways we could extend our study to improve the findings from it.
+In terms of the next steps, there are two ways we could extend our study to improve the findings from it.
 
 The first is by addressing the above mentioned limitation of combining data from various years to provide a more richer quality of identified results.
-By combining the data from past years, we are better able to learn of any temporal trends in the data which can lead to interesting results.
+By combining the data from past years, we are better able to learn of any temporal relationships in the data which can lead to stronger conclusions.
 
-The second way to improve our study is by making use of available streetcar delay sentiment data to understand the recurring pain points and emotions that are expressed by commuters.
-This data when combined with existing streetcar delay data can provide richer contextual information about the types of incident delays and the sentiment expressed.
-Another interesting result we can hope to learn is how these sentiments vary between various seasons as during harsher seasons like summer and winter, one can imagine that it wouldn't be most pleasant to experience a long delay in streetcar arrivals.
+The second way to improve our study is by making use of any available streetcar delay sentiment data to understand the recurring pain points and emotions that are expressed by commuters.
+This data when combined with existing streetcar delay data can provide richer contextual information about the types of incident delays and the sentiments expressed.
+Another interesting result we can hope to learn is how these sentiments vary between various seasons or times of day.
 
-These findings can help to better connect with the population and provide valuable information on how to better modify or establish new policies that can help to improve the overall commuting experience for the residents of Toronto, the city they call home.
+These findings can help TTC to better connect with the population and provide valuable information on how to improve policies and schedules that can help to improve the overall commuting experience for the residents of Toronto, the city they call home.
 

### 2024-09-27 paper/paper.qmd
Reupload file because they were added to the root

@@
 The following data was made available from OpenData Toronto (@opendatatoronto) for the year 2023.
-The data was extracted from the server and saved locally using in a raw data folder using the instructions provided by the documentation.
+The data was extracted from the server and saved locally in a raw data folder using the instructions provided by the documentation.
 We note that this was the only available dataset with frequent updates and data dating back to 2014.
@@
 The collected data represents common types like characters, numeric and datetime.
-Therefore, for the purposes of this study, the procedures used to collect the data and their types are appropriate for our use and we do not require any extra information to make any assumptions them.
+Therefore, for the purposes of this study, the procedures used to collect the data and their types are appropriate for our use and we do not require any extra information to make any assumptions about them.
 
